---
title: "CHORD: Classifier of HOmologous Recombination Deficiency"
output: rmarkdown::github_document
---

CHORD is a random forest model that uses the relative counts of somatic mutation contexts to predict
homologous recombination deficiency (HRD). The primary contexts used by CHORD are deletions with
flanking microhomology and 1-100kb structural duplications. Additionally, 1-100kb structural
duplications are used to distinguish BRCA1-type HRD from BRCA2-type HRD.

# Changes
CHORD has been updated to v2.0. In this version, the `del.mh` feature (deletions with microhomology)
is split into `del.mh.bimh.1` and `del.mh.bimh.2.5` (bases in microhomology: 1bp; 2 to >=5bp). This
was done to reduce false positive predictions. `del.mh.bimh.2.5` was found to be predictive of HRD
to a greater extent than `del.mh.bimh.1`. Furthermore, radiation therapy was found to be associated
with `del.mh.bimh.1`, which could result in false positive predictions for radiotherapy treated
patients.

# Reference
**Pan-cancer landscape of homologous recombination deficiency**  
*Luan Nguyen, John Martens, Arne Van Hoeck, Edwin Cuppen. Nat Commun 11, 5584 (2020).*
https://www.nature.com/articles/s41467-020-19406-4


# Installation
CHORD is itself an R package. Before using CHORD, some other R packages will also need to be 
installed, with the main one being mutSigExtractor, which is required for extracting the mutation 
contexts that CHORD uses. The below code can be used to install CHORD and mutSigExtractor, as well 
as the dependencies for both packages.

```{r eval=FALSE, error=FALSE}
## Bioconductor packages required by mutSigExtractor
install.packages('BiocManager')
BiocManager::install('BSgenome') ## Install genome parser
BiocManager::install('BSgenome.Hsapiens.UCSC.hg19') ## Install the default genome

## randomForest is required by CHORD
install.packages('randomForest')

## Install CHORD and mutSigExtractor directly from github using devtools
install.packages("devtools")
devtools::install_github('https://github.com/UMCUGenetics/mutSigExtractor/')
devtools::install_github('https://github.com/UMCUGenetics/CHORD/')
```

# Quick start
## Vcfs as input
CHORD can be supplied with vcf files containing **somatic (no germline)** SNVs, indels,
and SVs per sample. The first step is to extract the mutation contexts required by CHORD.
```{r eval=FALSE, error=FALSE}
contexts <- extractSigsChord(
  vcf.snv = '/path/to/vcf/with/snvs/',
  vcf.indel = '/path/to/vcf/with/indels/',
  vcf.sv = '/path/to/vcf/with/svs/',
  sv.caller = 'gridss' ## Can be 'gridss' (default) or 'manta'
)
```
Note that there is no real standard for SV vcfs. However, parsing vcfs produced by Manta or GRIDSS
has been built into CHORD.

## Dataframes as input
Alternatively, dataframes can be used as input (handy if you want to parse the vcfs yourself and 
supply this preprocessed data to CHORD).

```{r eval=FALSE, error=FALSE}
contexts <- extractSigsChord(
  df.snv = bed_file_like_dataframe_with_snvs,
  df.indel = bed_file_like_dataframe_with_indels,
  df.sv = dataframe_with_svtype_and_svlen
)
```

For SNVs and indels, a dataframe with the columns: chrom, pos, ref, alt.
```
##   CHROM      POS REF ALT
## 1     1 16145827   A   C
## 2     1 16492085   G   C
## 3     1 17890303   C   G
## 4     1 18877885   G   A
## 5     1 18919776   T   C
```

For SVs, a dataframe with the columns: sv_type, sv_len. For sv_type, values must be DEL, DUP, INV, 
TRA (deletions, duplications, inversions, translocations). For translocations, sv_len information is 
discarded.

```
##   sv_type    sv_len
## 1     TRA        NA
## 2     DEL      1696
## 3     DEL     22644
## 4     DUP      1703
## 5     DEL      1789
## 6     DEL     49256
```

The column names themselves do not matter, as long as the columns are in the aforementioned order.

## Inputs are flexible
The vcf.* and df.* arguments can be mixed. This is especially useful in the case of SV vcfs, since
it is possible that you will need to parse these yourself (in this case the ```sv.caller``` 
argument can be left out). Make sure to only use PASS variants! If your vcfs contain unfiltered 
variants, you can provide the desired FILTER values to keep to `vcf.filters`.
```{r eval=FALSE, error=FALSE}
contexts <- extractSigsChord(
  vcf.snv = '/path/to/vcf/with/snvs/',
  vcf.indel = '/path/to/vcf/with/indels/',
  df.sv = dataframe_with_svtype_and_svlen,
  
  ## This is optional (for when your vcfs are not filtered for PASS variants)
  vcf.filters=list(snv="PASS",indel="PASS",sv="PASS") 
)
```

Often, SNVs and indels are reported in the same vcfs. In such cases, the vcf path can be specified to
the `vcf.snv` argument (`vcf.indel` can be left out). Alternatively, SNVs and indels can be
provided as a dataframe, which can be specified to `df.snv` (with `df.indel` left out).
```{r eval=FALSE, error=FALSE}
contexts <- extractSigsChord(
  vcf.snv = '/path/to/vcf/with/snvs_and_indels/', ## or: df.snv = bed_file_like_dataframe_with_snvs_and_indels
  df.sv = dataframe_with_svtype_and_svlen
)
```

A different reference genome than the default (BSgenome.Hsapiens.UCSC.hg19) can be used. Genomes 
should be BSgenomes. The **variable name** (i.e. no quotes) of the BSgenome object is specified to
`ref.genome`.

```{r eval=FALSE, error=FALSE}
## Make sure to install and load the desired ref genome first
install.packages('BiocManager')
BiocManager::install('BSgenome.Hsapiens.UCSC.hg38')

## Non-default genomes need to be explicitly loaded. The default (BSgenome.Hsapiens.UCSC.hg19)
## is automatically loaded.
library(BSgenome.Hsapiens.UCSC.hg38)

## Specify the name of the BSgenome object to the ref.genome argument
contexts <- extractSigsChord(
  vcf.snv = '/path/to/vcf/with/snvs_and_indels/',
  df.sv = dataframe_with_svtype_and_svlen
  ref.genome=BSgenome.Hsapiens.UCSC.hg38
)
```


## Predicting HRD
Once we have the mutation contexts, a prediction can be made.
```
chordPredict(contexts)
```

# Tutorial
This tutorial will demonstrate how HRD prediction can be performed locally. For large vcfs and/or 
many samples, it is advised to run CHORD on a high-performance cluster (HPC).

Somatic vcfs from a few primary tumors from the 560 breast cancer cohort (BRCA-EU, ICGC) will be 
used as example input. These are located at `example/vcf/` in the CHORD git repository. These 
vcfs have been generated from tsv files which are publically available at https://dcc.icgc.org/repositories.

In part 1, we will use the vcfs directly as input for CHORD. 

Part 2 of this tutorial will provide a demonstration of how we can parse the vcfs into dataframes
that CHORD accepts. This is particularly useful if you have non-standard vcfs, as is often the 
case with SV vcfs. Of course, it is possible to produce the required input dataframes from
any data source, and not just vcfs; it only requires a bit of data wrangling :).

```{r message=FALSE, warning=FALSE, results="hide", echo=FALSE}
setwd('/Users/lnguyen/hpc/cuppen/projects/P0013_WGS_patterns_Diagn/CHORD/processed/scripts_main/CHORD/example/')
```

To run the example, clone/download the git repository to your own computer (e.g. to /Users/lnguyen/Desktop/), 
and set the working directory to the `example/` directory in the CHORD package.
```{r eval=FALSE}
setwd('/Users/lnguyen/Desktop/CHORD/example/')
```

## 1. Running CHORD on vcfs
### Loading vcfs into R
We will first create a dataframe to store the vcf paths as well as assign them sample names.

```{r message=FALSE, warning=FALSE}
library(CHORD)

options(stringsAsFactors=F)

## Get file paths from the vcf/ subdirectory
vcf_files <- data.frame(
  snv_indel=list.files('vcf', pattern='*snv_indel.vcf.gz', full.names=T),
  sv=list.files('vcf', pattern='*sv.vcf.gz', full.names=T)
)

## Assign sample names to files
vcf_files$sample <- sapply(strsplit(basename(vcf_files$snv_indel),'_'),`[`,1)

vcf_files
```

Vcfs can either be in plain text format (.vcf) or compressed (.gz).

SNVs and indels are often stored in the same vcf (as is the case with the example vcfs). CHORD
therefore accepts such vcfs as input; SNVs and indels are automatically split up. However, it is
also possible to specify the SNV and indel vcfs separately (by specifying the paths separately to
`vcf.snv` and `vcf.indel` to `extractSigsChord()`; see below).

### Extracting mutation contexts
Create a directory to write all subsequent output. **Note**: the output from this tutorial exists in 
the ```output/``` directory by default. Delete this directory to create new output files.
```{r message=FALSE, warning=FALSE}
dir.create('output/')
```

Extract the features that are used as input for CHORD. With the example data, it should take
about 30s to run.
```{r message=FALSE, warning=FALSE}
## Make dir to output contexts
contexts_dir <- 'output/contexts/'
dir.create(contexts_dir, recursive=T)

## Extract contexts for all samples
for(i in 1:nrow(vcf_files)){
  
  params <- as.list(vcf_files[i,])
  out_path <- paste0(contexts_dir,'/',params$sample,'_contexts.txt')
  
  if(!file.exists(out_path)){
    extractSigsChord(
      vcf.snv=params$snv_indel,
      vcf.sv=params$sv, sv.caller='manta',
      sample.name=params$sample,
      output.path=out_path, verbose=F
    )
  }
}
```
The `sv.caller` parameter has been included in `extractSigsChord()` since different SV
callers report SVs in different ways. `extractSigsChord()` includes parsing for vcfs produced by
Manta and GRIDSS (`sv.caller='manta'`or `sv.caller='gridss'`). If you have vcfs from other
SV callers, a solution is to provide the relevant data as a dataframe (see section 2 of the 
tutorial).

We can then merge the contexts into a matrix.
```{r message=FALSE, warning=FALSE}
## Get the paths of the contexts txt files
context_files <- list.files(contexts_dir, full.names=T)

## Read the txt files into R
l_contexts <- lapply(context_files, function(i){
  read.delim(i, check.names=F)
})

## Merged the contexts into a matrix
merged_contexts <- do.call(rbind, l_contexts)

## Write to output directory
write.table(merged_contexts, 'output/merged_contexts.txt', sep='\t', quote=F)
```

Below we can see what a part of the context matrix looks like.
```{r message=FALSE, warning=FALSE}
merged_contexts[,1:5]
```

### Predicting HRD and interpreting CHORD's output
Once we have the context matrix ready, we can use it for predicting HRD.
```{r message=FALSE, warning=FALSE}
chord_output <- chordPredict(merged_contexts, do.bootstrap=T, verbose=F)
write.table(chord_output, 'output/chord_pred.txt', sep='\t', quote=F)
```

#### Main output
To extract the HRD probabilities:
```{r message=FALSE, warning=FALSE}
chord_output[,1:8]
```

CHORD outputs the probability of:

* HRD: `p_hrd`
* BRCA1-type HRD: `p_BRCA1`
* BRCA2-type HRD: `p_BRCA2`

`hr_status` tells us if a sample is HR deficient (`p_hrd` >= 0.5) 
or proficient. If a sample is HRD, then `hrd_type` will tell us if the sample has BRCA1-type HRD 
or BRCA2-type HRD (=max(`p_BRCA1`,`p_BRCA2`)).

#### Pre-requisites for accurate HRD prediction
CHORD requires **>=50 indels** to accurately determine whether a sample is HRD. If this criterion is not
met, `hr_status` will be `cannot_be_determined` and `remarks_hr_status` will be 
`<50 indels`. 

CHORD **cannot be applied to MSI samples**. If an MSI sample is detected, `hr_status` will be 
`cannot_be_determined` and `remarks_hr_status` will be `Has MSI (>14000 indel.rep)`

If a sample is HRD, CHORD requires **>=30 SVs** to accurately determine HRD subtype. If this criterion is not
met, `hrd_type` will be `cannot_be_determined`, and `remarks_hr_status` will be 
`<30 SVs`.

The user may of course ignore these remarks and proceed with using the raw probabilities 
outputted by CHORD (`p_hrd` and/or `p_BRCA1`/`pBRCA2`).

#### Stability of predictions
To extract the bootstrap predictions:
```{r message=FALSE, warning=FALSE}
chord_output[,9:ncol(chord_output)]
```

To assess the stability of prediction for each sample, bootstrapping is performed by resampling the
feature vector 20 times and calculating HRD probabilities for each iteration. The probabilities at
the 5%, 50% 95% quantiles are then calculated. In other words, the bootstraped predictions provide 
a lower (5% quantile) and upper (95% quantile) error from the median probability (50% quantile) for 
each prediction class.

The user may choose to use the median probabilities in place of the single probabilities (as in
`chord_output$pred`). This can give more accurate HRD probabilities, especially for samples 
with low mutational load.

Note that performing the bootstrapping is computationally expensive. Bootstrapping can be turned off
by specifying `do.bootstrap=F` to `chordPredict()`.

## 2. Running CHORD from dataframes
To run CHORD on non-standard vcfs or from other sources, we can create dataframes that 
`extractSigsChord()` accepts. The output can then be passed to `chordPredict()`.

In this part of the tutorial, we will create the required dataframes from the example vcfs. This
will thus also serve as a short demonstration on how one could parse vcfs in R. Furthermore, we will 
focus on one sample for simplicity's sake.

### SNV and indel dataframes
For SNVs and indels, a dataframe containing CHROM, POS, REF and ALT information (i.e. a bed file like 
format) is required. The column names of this dataframe can be anything; as long as the columns are 
in the aforementioned order. 

We can use `readVcfFields()` from the `mutSigExtractor` package to read a vcf into R as a 
dataframe.

```{r message=FALSE, warning=FALSE}
library(mutSigExtractor)
```

```{r message=FALSE, warning=FALSE}
df_snv_indel <- readVcfFields('vcf/PD3905_snv_indel.vcf.gz', fields=c('CHROM','POS','REF','ALT'))
```

Alternatively, we can specify the vcf columns we want by index.
```{r eval=FALSE, error=FALSE}
df_snv_indel <- readVcfFields('vcf/PD3905_snv_indel.vcf.gz', fields=c(1,2,4,5))
```

The dataframe shown below is in the correct format as required by CHORD. This dataframe contains
both SNVs and indels. This is not a problem as we can specify the argument```df.snv``` (and leaving
out ```df.indel```) in ```extractSigsChord()``` as is done later in the tutorial.
```{r message=FALSE, warning=FALSE}
df_snv_indel[25:30,]
```


### Extracting SV contexts
For SVs, a dataframe is required with SV type (values must be one of the following: DEL,
DUP, INV, TRA) and SV length information. Again, the column names of this dataframe can be anything;
as long as the columns are in the aforementioned order.

Again, `readVcfFields()` can be used to read the vcf. The example vcfs have been produced by 
Manta, which reports SVTYPE and SVLEN in the INFO field, so only this field needs to be extracted 
from the vcf.

```{r message=FALSE, warning=FALSE}
vcf_sv <- readVcfFields('vcf/PD3905_sv.vcf.gz', fields='INFO')
print(vcf_sv[1:5,,drop=F], right=F)
```

Then, split the data each INFO entry, which are separated by ```;```.
```{r message=FALSE, warning=FALSE}
vcf_sv_info <- strsplit(vcf_sv$INFO,';')
vcf_sv_info[1:2]
```

Extract the SVTYPE and SVLEN data, and put it in the correct dataframe format.
```{r message=FALSE, warning=FALSE}
sv_type <- sapply(vcf_sv_info,`[`,2) ## Select the 2nd object from each INFO entry
sv_type <- gsub('SVTYPE=','',sv_type) ## Remove the 'SVTYPE=' prefix

sv_len <- sapply(vcf_sv_info,`[`,1) ## Select the 1st object from each INFO entry
sv_len <- gsub('SVLEN=','',sv_len) ## Remove the 'SVLEN=' prefix
sv_len <- as.integer(sv_len) ## Convert the character vector to an integer vector

df_sv <- data.frame(sv_type, sv_len)
head(df_sv)
```

You may notice that translocations (TRA) have an SV length, which doesn't really make sense. This is
no problem since SV length info is discarded for translocations.

### Predicting HRD
Extract the relevant mutation contexts from the variant data.
```{r message=FALSE, warning=FALSE}
contexts <- extractSigsChord(
  df.snv = df_snv_indel,
  df.sv = df_sv,
  sample.name='PD3905'
)

```

Below we can see what a part of the context matrix looks like.
```{r message=FALSE, warning=FALSE}
contexts[,c(1:4, 97:104, 127:131),drop=F]
```

Lastly, make the HRD prediction.
```{r message=FALSE, warning=FALSE}
chord_output <- chordPredict(contexts, verbose=F)
chord_output
```

Please refer back to section 1 of the tutorial for interpreting CHORD's output.

```{r error=FALSE, results="hide", echo=FALSE}
pkg_dir  <- '/Users/lnguyen/hpc/cog_bioinf/cuppen/project_data/Luan_projects/CHORD/scripts_main/CHORD/'
file.copy(
   paste0(pkg_dir,'/example/README.md'),
   paste0(pkg_dir,'/README.md'),
   overwrite=T
)
```

